---
title: "Compulsory Exercise 2: Prediction of Alzheimers Disease from Longitudinal MRI Data"
author:
- Chester Henry Charlton
- Kasper Eikeland
- Tinus Garshol
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes: \usepackage{amsmath}
output:
  # html_document:
  #   toc: no
  #   toc_depth: '2'
  #   df_print: paged
  pdf_document:
    toc: no
    toc_depth: '2'
urlcolor: blue
abstract: "This is the place for your abstract (max 350 words)"
---
  
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,message=FALSE,warning=FALSE,strip.white=TRUE,prompt=FALSE,
                      cache=TRUE, size="scriptsize",fig.width=4, fig.height=3,fig.align = "center")

```

```{r,eval=TRUE,echo=FALSE}
library("knitr")
library("rmarkdown")
library(GGally)
```

```{r}
setwd("/Users/chcharlton/NTNU/statLearning/tma4268_alzheimers")
```


```{r}
alzheimers.df <- read.csv("oasis_longitudinal.csv")
alzheimers.df$Group <-as.factor(alzheimers.df$Group)
alzheimers.df.working <- alzheimers.df[,-c(2,)]
tail(alzheimers.df,30)
```

<!--  Henry's initial work here -->
NB: here is the link to the article describing this data in detail: https://direct.mit.edu/jocn/article/22/12/2677/4983/Open-Access-Series-of-Imaging-Studies-Longitudinal
how about adding some CV plots, Cp, BIC, something like that. 

```{r}
# select some quantitative data and plot the pairs of it
pairwise_scatter_data <- alzheimers.df[,c("Age", "EDUC", "MMSE", 
                                   "eTIV", "nWBV", "ASF")]
ggpairs(data=pairwise_scatter_data)
```
```{r}
# make some box plots on CDR
par(mfrow=c(2,4))  
boxplot(Age~CDR,data=alzheimers.df)
boxplot(EDUC~CDR,data=alzheimers.df)
boxplot(SES~CDR,data=alzheimers.df)
boxplot(MMSE~CDR,data=alzheimers.df)
boxplot(eTIV~CDR,data=alzheimers.df)
boxplot(nWBV~CDR,data=alzheimers.df)
boxplot(ASF~CDR,data=alzheimers.df)

# make some box plots on Group
par(mfrow=c(2,3))   
boxplot(Age~Group,data=alzheimers.df)
boxplot(EDUC~Group,data=alzheimers.df)
boxplot(MMSE~Group,data=alzheimers.df)
boxplot(eTIV~Group,data=alzheimers.df)
boxplot(nWBV~Group,data=alzheimers.df)
boxplot(ASF~Group,data=alzheimers.df)

```







<!--  Kasper's initial work here. -->


<!--  Tinus' initial work here. -->


## Introduction: Scope and purpose of your project


## Descriptive data analysis/statistics
### Description of Data Types:
Group: The group that the patient was classified in at the across the various scans. Possible values are "Undemented," denoting a patient without symptoms of dementia throughout all scans, "Demented" denoting a patient with symptoms of dementia accross all scans, or "Converted," denoting someone who began without symptoms of dementia but developed them at a further scan. 
Visit: The order of the visit in which the scan was done, indexed from 1. 
MR Delay: The number of days since the previous scan. Marked as 0 for the initial scan.
M.F: Categorical variable for male or female patients at birth. M denotes male, F denotes female. 
Hand: The patient's dominant hand. All of the patients were right-handed. 
EDUC: The education level of the patient, measured in years (including elementary school). 
SES: "Socioeconomic status as assessed by the Hollingshead Index of Social Position and classified into categories from 1 (highest status) to 5 (lowest status) (Hollingshead, 1957)"
MMSE: "Mini-Mental State Examination score (range is from 0 = worst to 30 = best) (Folstein, Folstein, & McHugh, 1975)"
CDR: "Clinical Dementia Rating (0 = no dementia, 0.5 = very mild AD, 1 = mild AD, 2 = moderate AD) (Morris, 1993)"
ASF: "Atlas scaling factor (unitless). Computed scaling factor that transforms native-space brain and skull to the atlas target (i.e., the determinant of the transform matrix) (Buckner et al., 2004)"
eTIV: "Estimated total intracranial volume (cm3) (Buckner et al., 2004)"
nWBV: "Normalized whole-brain volume, expressed as a percent of all voxels in the atlas-masked image that are labeled as gray or white matter by the automated tissue segmentation process (Fotenos et al., 2005)"



## Methods


## Results and interpretation


## Summary
